Rates and delay times of type Ia supernovae in the Dark Energy Survey

### 2021-03-08 draft.tex
Updates from Overleaf

% The list of authors, and the short list which is used in the headers.
% If you need two or more lines of authors, add an extra line using \newauthor
\author[P. Wiseman et al.]{
P. Wiseman,$^{1}$\thanks{E-mail: p.s.wiseman@soton.ac.uk}
A. N. Other,$^{2}$
Third Author$^{2,3}$
and Fourth Author$^{3}$
\\
}

@@

% Abstract of the paper
\begin{abstract}
We use a sample of 881 photometrically classified type Ia supernovae (SNe Ia) discovered by the Dark Energy Survey (DES) in the redshift range $0.2 < z <0.6$, along with 40415 field galaxies to calculate the rate of SNe Ia per galaxy. We recover the established positive trend of SN Ia rate as a function of galaxy stellar mass across a broad range of scales $8.5 \leq \log(M_*/\mathrm{M}_{\odot}) \leq 11.25$. We find that the SN Ia rate increases with stellar mass roughly linearly in log space, with a slope of $0.63 \pm 0.02$, which is consistent with previous work. We use an empirical model of stellar mass assembly to estimate the average star-formation histories (SFHs) of galaxies across the stellar mass range of our measurement. Combining the modelled SFHs with the SN Ia rates to estimate constraints on the SN Ia delay time distribution (DTD), we find the data are best fit by a power-law DTD with index $\beta = -1.14 \pm 0.05$ and normalisation $A = 2.1 \pm0.05 \times 10^{-13}~\mathrm{SNe}~{\mathrm{M}_{\odot}}^{-1}~\mathrm{yr}^{-1}$. Upon splitting the SN sample by properties of the light curves, we find a strong dependence on DTD slope with the SN stretch parameter, with slower-declining SNe exhibiting a steeper DTD slope. We interpret this as a result of a relationship between stretch and progenitor age, and explore the inplications of the result in the context of SN Ia progenitors.
\end{abstract}

@@

Type Ia supernovae (SNe Ia) are explosions of white dwarf stars (WDs) at masses close to the Chandrasekhar limit of 1.44 $M_{\odot}$ ($M_{\mathrm{Ch}}$. The small observed dispersion in their peak brightnesses can be reduced further through empirical relations between brightness and light curve properties, such as decline rate (stretch) or optical colour \citep{Phillips1993,Tripp1998}. These properties have led SNe Ia to be used extensively by cosmologists to measure relative distances in the Universe \citep{Perlmutter1998,Riess1998}

Despite using these relations to correct SN Ia peak brightnesses to within a dispersion of $\sim 0.1$ mag in samples with over one thousand SNe \citep{Scolnic2018}, the exact nature of the progenitor scenarios for SNe Ia is yet to be confirmed. While it is accepted that SNe Ia are caused by mass transfer onto a WD from a companion star, multiple possible scenarios exist for the nature of that companion star (see \citealt{Maoz2014} for a review). The leading models involve a main sequence (MS) star in the single-degenerate (SD) scenario \citep{Whelan1973,Nomoto1982}, a secondary white dwarf in the double-degenerate (DD) scenario \citep{Tutukov1976,Iben1984,Webbink1984}, while more exotic models such as the core-degenerate (CD) scenario invoke white dwarves merging with the cores of massive stars \citep{Soker2019}.

The progenitors of a number of core-collapse SNe (CCSNe) have been identified in high-resolution pre-explosion images, allowing detailed analysis of the latter stages of massive star evolution. However, this approach has not yet been successful in identifying a progenitor of a SN Ia. Conversely, it has been possible to search for surviving remnants of the binary system -- main sequence stars or white dwarves left behind at the SN location or kicked \citep{Shen2018} -- displaying tentative evidence for a DD scenario. Without an observation of a SN Ia progenitor system or remnant companion, there is as yet no direct evidence ruling out any progenitor channel from being a significant contribution to the total population of SNe Ia.

@@

\label{sec:data}
\subsection{Dark Energy Survey supernova programme \label{subsec:des}}
The DES Supernova Programme (DES-SN) is a transient survey based on five six-month seasons of observations with the Dark Energy Camera (DECam; \citealt{Flaugher2015}), designed primarily to measure the light curves of SNe Ia for use as cosmological distance indicators. Transients were detected and processed using a difference imaging program \citep{Kessler2015} and vetted using a machine learning (ML) algorithm \citep{Goldstein2015}. Spectral follow-up was performed on a suite of large optical telescopes \citet{Smith2020a}, leading to the initial publication of a measurement of cosmological parameters using 207 spectroscopically confirmed SNe (\citealt{DESCollaboration2018a}, and references therein). 
\subsection{Supernovae \label{subsec:host_sample}}
\subsubsection{Photometric classification and quality cuts \label{subsubsec:sn_classify}}
With over 30,000 transients passing ML cuts it was not possible to obtain spectral follow up of every object. Instead we make use of photometric classification in the form of the recurrent neural network classifier \texttt{superNNova} (SNN; \citealt{Moller2019}), mirroring the approach of \citet{Scolnic2020}. We train SNN using a large suite of simulated multi-band SN light curves. \textbf{Waiting on final sample and a few words from Maria/Anais to complete this bit.} SNe Ia are simulated based on the \texttt{SALT2} model \citep{Guy2007} trained on the Joint Lightcurve Analysis (JLA) data set \citep{Betoule2014}. CCSNe are generated from the templates of \citet{Vincenzi2019}. We select SNe with a threshold probability of $P(\mathrm{Ia})\geq0.5$, although the results of this analysis are insensitive to the choice of threshold, since the vast majority of SNe receive classifications close to 1 or 0. The photometrically classified SNe Ia are then passed through the \texttt{SALT2} framework light curve fitter and are subject to quality cuts on the light curve width (stretch; $x_1$) and colour ($c$) parameters in an identical manner to \citet{Vincenzi2020}: $-3 \leq x_1 \leq 3$, $-0.3 \leq c \leq 0.3$. These cuts help to reduce the potential contamination from core-collapse SNe (CCSNe) and outlying thermonuclear SNe. We impose further cuts on the quality of those measured parameters, which are: the uncertainty on $x_1$: $\sigma_{x_1} <1$; the uncertainty on the date at which the light curve peaked: $\sigma_{t_{\mathrm{peak}}} <2$; the SALT2 fit probability $>0.001$.

@@

The numbers of SN hosts and field galaxies passing these cuts are listed in Table \ref{tab:cuts}. The final samples comprise 881 SNe and their host galaxies and 48,177 field galaxies. The volume-weighted mean redshifts are 0.50 for both SNe and field galaxies.

@@

We estimate global galaxy properties for both the SN host and field galaxies that pass the quality cuts by fitting the photometry with stellar population templates in a method outlined by \citet{Sullivan2006} and consistent to that used in previous DES-SN analyses \citep{Kelsey2020,Smith2020,Wiseman2020}. For SN hosts the redshift is fixed at the spectroscopically determined value, while for field galaxies we fix it at either the spectroscopic value if one exists in the OzDES GRC, or more commonly the photometrically derived value. We use the stellar population templates of \citet{Bruzual2003} and adopt a \citet{Chabrier2003} initial mass function (IMF). The fitting procedure returns a best fitting template and corresponding stellar mass ($M_*$). Upper and lower bounds on stellar mass estimates are taken as the largest and smallest values of a given parameter for which the reduced $\chi^2$ parameter corresponds to $\chi^2_{\mathrm{red, min}} +2.3$. To check that the choice of templates do not bias the result, we also fit the galaxies using the P\'EGASE.2 templates \citep{Fioc1997,LeBorgne2002} and a \citet{Kroupa2001} IMF, and find results consistent within measurement uncertainties. This is consistent with the findings of \citet{Smith2020}.

@@

By integrating the normalised DTD over cosmic time, we obtain an estimate of the average SN Ia efficiency $N_{\mathrm{Ia}}/M_*$ which represents the number of SNe Ia formed per unit mass of stars formed. We measure $N_{\mathrm{Ia}}/M_* = 0.9~_{-0.7}^{+4.0} \times 10^{-3}~\mathrm{SNe}~\mathrm{M}_{\odot}^{-1}$, which is consistent with \citet{Graur2011}, \citet{Maoz2011}, \citet{Perrets2012}, and \citet{Graur2013}

@@

\begin{figure}
    \centering
    \includegraphics[width=.5\textwidth]{figs/rate_vs_mass_DTD_fit_beta_norm_Qerf1.1_split_x1.png}
    \caption{As per Fig. \ref{fig:rate_fitted} but with the data split by their stretch parameter $x_1$.%The blue lines indicate the best fitting model based on a $t^{\beta}$ power law DTD convolved with SFHs based on the \citet{Childress2014} model}
    \label{fig:rate_fitted_split_x1}}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=.5\textwidth]{figs/rate_vs_mass_DTD_fit_beta_norm_Qerf1.1_split_x1_fitjoint.png}
    \caption{As per Fig. \ref{fig:rate_fitted_split_x1} but with the data fit by a two-component DTD.%The blue lines indicate the best fitting model based on a $t^{\beta}$ power law DTD convolved with SFHs based on the \citet{Childress2014} model}
    \label{fig:rate_fitted_split_x1_2c_DTD}}
\end{figure}
The SN Ia rate as a function of stellar mass SNe split by stretch is shown in Fig. \ref{fig:rate_fitted_split_x1}. The evolution of the rate with stellar mass is significantly different for the two sub-samples: high-stretch SNe dominate the rate at low stellar mass but tail off in the higher mass galaxies, while low-stretch SNe are subdominant in low mass galaxies but display a much steeper dependence on stellar mass and make up the vast majority of SNe in high-mass hosts. These measurements reflect the observed correlations of $x_1$ with stellar mass and sSFR (and by inference, stellar age) as been seen by \citet{Rigault2013,Rigault2018,Nicolas2020}. The corresponding DTD fit results in a steep power-law $\beta = -1.68 \pm 0.15$ for high-stretch SNe indicative of a population of predominantly prompt SNe, and a much shallower decay slope ($\beta = -0.79 \pm 0.08$) for low-stretch SNe representing a much more delayed population.

@@

One possibility is that sub-luminous SNe are more numerous in such predominantly old stellar populations, such that the fraction passing our light curve cuts is lower than the rest of the sample. Such a phenomenon could be caused by SN Ia sub-classes such as SN 1991bg-like SNe, which are known to explode exclusively in old environments \citep{Perets2010} at delay times as long as $>6$~Gyr \citep{Panther2019}. If the SN 1991bg-like objects are drawn from the same parent population of WDs as the "normal" SNe Ia, they could explain the lack thereof in the sample.

@@

The predictions from the DTD model diverge from the data at low stellar mass ($M_* <10^{9.25}$ M$_{\odot}$), where the rate becomes dominated by blue SNe. Although this is where the data is noisiest, it is worth investigating the possible causes of this divergence. A slight preference for SN to be bluer in lower-mass galaxies has been observed \citep{Scolnic2018,Brout2019,Smith2020,Kelsey2020} and could be the same effect seen at low stellar masses here. Physically, the dominance of blue SNe at low stellar masses could either be an intrinsic property of the SNe such as a hotter photosphere in explosions of younger, more massive progenitors, or an extrinsic effect such as a lower average dust column density in low-mass host galaxies. An intermediate explanation is that the SNe in low-mass hosts have a smaller dust column density in the immediate vicinity of the progenitor system, either due to the shorter evolutionary timescale or the higher mass and thus brighter system being less conducive to dust production. A detailed exploration of the effects of environment on SN colour will be presented in Kelsey et al. \textit{in prep}.

>>>>

% The list of authors, and the short list which is used in the headers.
% If you need two or more lines of authors, add an extra line using \newauthor
\author[P. Wiseman et al.]{
P. Wiseman,$^{1}$\thanks{E-mail: p.s.wiseman@soton.ac.uk}
M. Sullivan,$^{1}$
C. Frohmaier,$^{2,1}$
M. Smith, $^{3}$
M. Vincenzi, $^{2}$
et al.
\\
}

@@

% Abstract of the paper
\begin{abstract}
We use a sample of 881 photometrically classified type Ia supernovae (SNe Ia) discovered by the Dark Energy Survey (DES) in the redshift range $0.2 < z <0.6$, along with 40415 field galaxies to calculate the rate of SNe Ia per galaxy. We recover the established positive trend of SN Ia rate as a function of galaxy stellar mass across a broad range of scales $8.5 \leq \log(M_*/\mathrm{M}_{\odot}) \leq 11.25$. We find that the SN Ia rate increases with stellar mass roughly linearly in log space, with a slope of $0.63 \pm 0.02$, which is consistent with previous work. We use an empirical model of stellar mass assembly to estimate the average star-formation histories (SFHs) of galaxies across the stellar mass range of our measurement. Combining the modelled SFHs with the SN Ia rates to estimate constraints on the SN Ia delay time distribution (DTD), we find the data are fit well by a power-law DTD with index $\beta = -1.14 \pm 0.05$ and normalisation $A = 2.1 \pm0.05 \times 10^{-13}~\mathrm{SNe}~{\mathrm{M}_{\odot}}^{-1}~\mathrm{yr}^{-1}$. Upon splitting the SN sample by properties of the light curves, we find a strong dependence on DTD slope with the SN stretch parameter, with slower-declining SNe exhibiting a steeper DTD slope. We interpret this as a result of a relationship between stretch and progenitor age, and explore the implications of the result in the context of SN Ia progenitors.
\end{abstract}

@@

Type Ia supernovae (SNe Ia) are explosions of white dwarf stars (WDs). Although SNe Ia show diversity in their observed properties, a large fraction of them display a small dispersion in their peak brightnesses which can be reduced further through empirical relations between brightness and light curve properties, such as decline rate (stretch) or optical colour \citep{Rust1974,Pskovskii1977,Phillips1993,Tripp1998}. These properties have led SNe Ia to be used extensively by cosmologists to measure relative distances in the Universe \citep{Riess1998,Perlmutter1999}.

Despite using these relations to correct SN Ia peak brightnesses to within a dispersion of $\sim 0.1$ mag in samples with over one thousand SNe \citep{Scolnic2018}, the exact nature of the progenitors of SNe Ia is yet to be confirmed. While it is likely that SNe Ia are caused by mass transfer onto a WD from, or violent merger with, a companion star, multiple possible scenarios exist for the nature of that companion star (see \citealt{Maoz2014} for a review). The leading models involve a main sequence (MS) star in the single-degenerate (SD) scenario \citep{Whelan1973,Nomoto1982}, a secondary white dwarf in the double-degenerate (DD) scenario \citep{Tutukov1976,Iben1984,Webbink1984}, or more exotic models such as the core-degenerate (CD) scenario which invoke WDs merging with the cores of massive stars \citep{Soker2019}.

The progenitors of a number of core-collapse SNe (CCSNe) have been identified in high-resolution pre-explosion images \citep{Smartt2009,Eldridge2013}, allowing detailed analysis of the latter stages of massive star evolution. However, this approach has not yet been successful in identifying a progenitor of a SN Ia (although see \citet{McCully2014} for evidence of a progenitor for a type Iax SN). Conversely, it has been possible to search for surviving remnants of the binary system -- main sequence stars or white dwarves left behind at the SN location or kicked \citep{Shen2018} -- displaying tentative evidence for a DD scenario. Without an observation of a SN Ia progenitor system or remnant companion, there is as yet no direct evidence ruling out any progenitor channel from being a significant contribution to the total population of SNe Ia.

@@

\label{sec:data}
\subsection{Dark Energy Survey supernova programme \label{subsec:des}}
The DES Supernova Programme (DES-SN) is a transient survey based on five six-month seasons of observations with the Dark Energy Camera (DECam; \citealt{Flaugher2015}), designed primarily to measure the light curves of SNe Ia for use as cosmological distance indicators. Transients were detected and processed using a difference imaging program \citep{Kessler2015} and vetted using a machine learning (ML) algorithm \citep{Goldstein2015}. Spectral follow-up of live SN candidates was performed on a suite of large optical telescopes \citet{Smith2020a}, leading to the initial publication of a measurement of cosmological parameters using 207 spectroscopically confirmed SNe Ia (\citealt{DESCollaboration2018a}, and references therein). 
\subsection{Supernovae \label{subsec:host_sample}}
\subsubsection{Photometric classification and quality cuts \label{subsubsec:sn_classify}}
With over 30,000 transients passing ML selection it was not possible to obtain spectral follow up of every object. Instead we make use of photometric classification in the form of the recurrent neural network classifier \texttt{superNNova} (SNN; \citealt{Moller2019}), mirroring the approach of \citet{Scolnic2020}. We train SNN using a large suite of simulated multi-band SN light curves. \textbf{Waiting on final sample and a few words from Maria/Anais to complete this bit.} SNe Ia are simulated based on the \texttt{SALT2} model \citep{Guy2007} trained on the Joint Lightcurve Analysis (JLA) data set \citep{Betoule2014}. CCSNe are generated from the templates of \citet{Vincenzi2019}. We select SNe with a threshold probability of $P(\mathrm{Ia})\geq0.5$, although the our analysis is not sensitive to this choice, since the vast majority of SNe receive classifications close to 1 or 0. The photometrically classified SNe Ia are then passed through the \texttt{SALT2} framework light curve fitter and are subject to quality cuts on the light curve width (stretch; $x_1$) and colour ($c$) parameters in an identical manner to \citet{Vincenzi2020}: $-3 \leq x_1 \leq 3$, $-0.3 \leq c \leq 0.3$. This selection helps to reduce the potential contamination from core-collapse SNe (CCSNe) and outlying thermonuclear SNe. We impose further cuts on the quality of those measured parameters, which are: the uncertainty on $x_1$: $\sigma_{x_1} <1$; the uncertainty on the date at which the light curve peaked: $\sigma_{t_{\mathrm{peak}}} <2$; the SALT2 fit probability $>0.001$.

@@

The numbers of SN hosts and field galaxies passing these cuts are listed in Table \ref{tab:cuts}. The final samples comprise 881 SNe and their host galaxies and 40,415 field galaxies. The volume-weighted mean redshifts are 0.50 for both SNe and field galaxies.

@@

We estimate global galaxy properties for both the SN host and field galaxies that pass the quality cuts by fitting the photometry with stellar population templates in a method outlined by \citet{Sullivan2006} and consistent to that used in previous DES-SN analyses \citep{kelsey2021,Smith2020,Wiseman2020}. For SN hosts the redshift is fixed at the spectroscopically determined value, while for field galaxies we fix it at either the spectroscopic value if one exists in the OzDES GRC, or more commonly the photometrically derived value. We use the stellar population templates of \citet{Bruzual2003} and adopt a \citet{Chabrier2003} initial mass function (IMF). The fitting procedure returns a best fitting template and corresponding stellar mass ($M_*$). Upper and lower bounds on stellar mass estimates are taken as the largest and smallest values of a given parameter for which the reduced $\chi^2$ parameter corresponds to $\chi^2_{\mathrm{red, min}} +2.3$. To check that the choice of templates do not bias the result, we also fit the galaxies using the P\'EGASE.2 templates \citep{Fioc1997,LeBorgne2002} and a \citet{Kroupa2001} IMF, and find results consistent within measurement uncertainties. This is consistent with the findings of \citet{Smith2020}.

@@

By integrating the normalised DTD over cosmic time, we obtain an estimate of the average SN Ia efficiency $N_{\mathrm{Ia}}/M_*$ which represents the number of SNe Ia formed per unit mass of stars formed. We measure $N_{\mathrm{Ia}}/M_* = 0.9~_{-0.7}^{+4.0} \times 10^{-3}~\mathrm{SNe}~\mathrm{M}_{\odot}^{-1}$, which is consistent with \citet{Graur2011}, \citet{Maoz2011}, \citet{perrett2012}, and \citet{Graur2013}

@@

\begin{figure}
    \centering
    \includegraphics[width=.5\textwidth]{figs/rate_vs_mass_DTD_fit_beta_norm_Qerf1.1_split_x1.png}
    \caption{As per Fig. \ref{fig:rate_fitted} but with the data split by their stretch parameter $x_1$. Fits were performed with the DTD slope $\beta$ and normalisation $A$ as free parameters, and prompt time $t_{\mathrm{p}}$ fixed at 0.04 Gyr.%The blue lines indicate the best fitting model based on a $t^{\beta}$ power law DTD convolved with SFHs based on the \citet{Childress2014} model}
    \label{fig:rate_fitted_split_x1}}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=.5\textwidth]{figs/rate_vs_mass_DTD_fit_beta_norm_Qerf1.1_split_x1_fitjoint.png}
    \caption{As per Fig. \ref{fig:rate_fitted_split_x1} but with the data fit simultaneously by a single two-component DTD, with fixed slope $\beta = -1$ and a prompt fraction that is 1 at short delay times, 0 at large delay times, and a linear function of time between $t_1$ and $t_2$.%The blue lines indicate the best fitting model based on a $t^{\beta}$ power law DTD convolved with SFHs based on the \citet{Childress2014} model}
    \label{fig:rate_fitted_split_x1_2c_DTD}}
\end{figure}
The SN Ia rate as a function of stellar mass SNe split by stretch is shown in Fig. \ref{fig:rate_fitted_split_x1}. The evolution of the rate with stellar mass is significantly different for the two sub-samples: high-stretch SNe dominate the rate at low stellar mass but tail off in the higher mass galaxies, while low-stretch SNe are subdominant in low mass galaxies but display a much steeper dependence on stellar mass and make up the vast majority of SNe in high-mass hosts. These measurements reflect the observed correlations of $x_1$ with stellar mass and sSFR (and by inference, stellar age) as been seen by \citet{Rigault2013,Rigault2018,Nicolas2020}. The corresponding DTD fit results in a steep power-law $\beta = -1.68 \pm 0.15$ for high-stretch SNe indicative of a population of predominantly prompt SNe, and a much shallower decay slope ($\beta = -0.79 \pm 0.08$) for low-stretch SNe representing a much more delayed population.

@@

One possibility is that sub-luminous SNe are more numerous in such predominantly old stellar populations, such that the fraction passing our light curve cuts is lower than the rest of the sample. Such a phenomenon could be caused by SN Ia sub-classes such as SN 1991bg-like SNe, which are known to explode exclusively in old environments \citep{Perets2010} at delay times as long as $>6$~Gyr \citep{Panther2019}. If the SN 1991bg-like objects are drawn from the same parent population of WDs as the "normal" SNe Ia, they could explain the lack thereof in the sample. We check this briefly by re-examining the objects left out of our sample due to \texttt{SALT2} cuts. We find that 75\% of objects left out lie in hosts with $M_*>10^{10}$ M$_{\odot}$, while this percentage is 65\% for the objects that pass the cuts. However, of the objects in high-mass galaxies that fail the cuts, only 11 (10\%) fail due to their $x_1$ lying below $-3$, and a total of 20 have $x_1 <-2$ which is the regime populated by SN 1991bg-like objects which is well short of the factor of two by which the DTD model over-predicts the data in the highest mass bin.

@@

The predictions from the DTD model diverge from the data at low stellar mass ($M_* <10^{9.25}$ M$_{\odot}$), where the rate becomes dominated by blue SNe. Although this is where the data is noisiest, it is worth investigating the possible causes of this divergence. A slight preference for SN to be bluer in lower-mass galaxies has been observed \citep{Scolnic2018,Brout2019,Smith2020,kelsey2021} and could be the same effect seen at low stellar masses here. Physically, the dominance of blue SNe at low stellar masses could either be an intrinsic property of the SNe such as a hotter photosphere in explosions of younger, more massive progenitors, or an extrinsic effect such as a lower average dust column density in low-mass host galaxies. An intermediate explanation is that the SNe in low-mass hosts have a smaller dust column density in the immediate vicinity of the progenitor system, either due to the shorter evolutionary timescale or the higher mass and thus brighter system being less conducive to dust production. A detailed exploration of the effects of environment on SN colour will be presented in Kelsey et al. \textit{in prep}.
